Manuscript Figures

### 2021-06-10 src/manuscript_figures.Rmd
update immunothearpy response figure with benchmarks included

dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]

top_vars <- unique(dt$variable)

# quantify prediction accuracy for finding complete/partial responders
m <- M[,top_vars]
f <- colData$binaryResponse
m <- m[!is.na(f),]
f <- f[!is.na(f)]

# heatmap of top vars 
p1 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)

# survival stratification using the top LF 
require(survminer)
require(survival)
top_lf <- dt[order(padj)][ref_cl == 'CR/PR'][1,]$variable
cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')

p2 <- plot_survival(surv = colData, partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 'high', 'low'), 
                                                        row.names = rownames(m)), 
              duration_col = 'os', event_col = 'censOS')

# top go terms for top features of top LF 
feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
# top pathways for top lf
top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:10]
top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 32)
p3 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
  geom_bar(stat = 'identity') + 
  coord_flip() + labs(x = '') 
colnames(feats)[2] <- 'top_lf'
p4 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
  geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')

# correlation of top LF with tumor mutation burden
# get samples with both response and tmb annotation
cd <- colData #colData[!is.na(`Neoantigen burden per MB`)][!is.na(binaryResponse)]
m2 <- M[cd$bcr_patient_barcode,]
require(ggpubr)
df <- data.frame('response' = cd$binaryResponse, 
                 'TMB' = cd$`Neoantigen burden per MB`, 
                 'LF' = m2[,top_lf])
p5 <- ggpubr::ggscatter(df, x = 'LF', y = 'TMB', 
                        add = 'reg.line', cor.coef = T, cor.coef.coord = c(30, 10), 
                  cor.coef.size = 4) + labs(x = top_lf, y = 'Tumor Mutation Burden')

# immune phenotypes 
ip <- colData$`Immune phenotype`
#ip[ip == 'excluded'] <- NA
m3 <- M[!is.na(ip),]
ip <- ip[!is.na(ip)]

dt3 <- get_factor_specific_variables(m3, ip)[order(pval)][padj < 0.01]
ip_vars <- unique(dt3$variable)
ip_vars <- get_top_by_label(dt3, 'ref_cl', topN = 5)
res_ip <- predict_cluster_labels_glmnet(m3[,unique(dt3$variable)], ip, returnROC = T)
p6 <- plot_roc_list(res_ip)

pheatmap::pheatmap(get_basis_matrix(m3[,unique(ip_vars$variable)], ip), silent = T)

Fig <- cowplot::plot_grid(cowplot::plot_grid(p1, p2$gtable, p4, p3$plot, 
                                             labels = 'AUTO'), 
                          cowplot::plot_grid(p5, p6, p7, nrow = 1, 
                                             labels = c('E', 'F', 'G')), 
                          nrow = 2, rel_heights = c(2, 1)) 
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.immunotherapy_response.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 10, height = 14)
Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
Cancers (N=348) using MAUI. A) AUC curve for predicting Complete/Partial Responders (CR/PR) to the anti-PD-L1
treatment using an Elastic Net Model on a subset of MAUI latent factors individually predictive of treatment response groups. B) Average activation score of top latent factors predictive of treatment response groups. C) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44). D) Survival stratification of the 
cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4). E) Top 10 genes contributing to LF44
F) Correlation between LF44 activation score and the tumor (neo-antigen) mutation burden"
```

>>>>

dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')]
dt$tool <- toupper(gsub("_factors", "", dt$tool))
dt$ref_cl <- ifelse(dt$ref_cl == 'CR/PR', 'Responders', 'Non-responders')

p1 <- ggplot(dt[,.SD[which.min(padj)],by = c('tool', 'ref_cl')], 
       aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5))

# heatmap of top vars 
# plot maui factors for responders/non-responders
M <- LFs_imtrp$maui_factors$mariathasan.gex
top_vars <- unique(dt[tool == 'MAUI'][padj < 0.01]$variable)
m <- M[,top_vars]
f <- ifelse(colData$binaryResponse == 'CR/PR', 'Responders', 'Non-responders')
m <- m[!is.na(f),]
f <- f[!is.na(f)]
p2 <- pheatmap::pheatmap(get_basis_matrix(m[,top_vars], f), silent = T)

# survival stratification using the top LF 
require(survminer)
require(survival)
top_lf <- dt[order(padj)][ref_cl == 'Responders'][1,]$variable
cp <- find_survival_cutpoint(m[,top_lf,drop = F], colData, event_col = 'censOS', duration_col = 'os')
p3 <- plot_survival(surv = colData, 
                    partition.df = data.frame('group' = ifelse(m[,top_lf] > cp, 
                                              paste0(top_lf, ">", round(cp, 1)),
                                              paste0(top_lf, "<=", round(cp, 1))), 
                                              row.names = rownames(m)), 
                    duration_col = 'os', event_col = 'censOS')

# top go terms for top features of top LF 
W <- FWs_imtrp$maui_factors$mariathasan.gex
feats <- data.table(W[,top_lf,drop=F], keep.rownames = T)[order(abs(get(top_lf)), decreasing = T)][1:100]
# top pathways for top lf
top_terms <- compute_geneset_enrichment(feats$rn, pathways)[order(pval)][padj < 0.05][1:5]
top_terms$term_wr <- stringr::str_wrap(gsub("_", " ", top_terms$term), width = 24)
p4 <- ggplot(top_terms, aes(x = reorder(term_wr, -pval), y = -log10(padj))) + 
  geom_bar(stat = 'identity') + 
  coord_flip() + labs(x = '') 
colnames(feats)[2] <- 'top_lf'
p5 <- ggplot(feats[1:10,], aes(x = reorder(rn, abs(top_lf)), y = abs(top_lf))) + 
  geom_bar(stat = 'identity') + coord_flip() + labs(x = '', y = 'Feature Importance (absolute)')

# correlation of top LF with tumor mutation burden
# get samples with both response and tmb annotation
dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
  M <- LFs_imtrp[[tool]]$mariathasan.gex
  dt <- data.table(cor(M, colData$`Neoantigen burden per MB`, use = 'complete.obs'),
                   keep.rownames = T)
  dt$tool <- tool
  return(dt)
}))
dt <- dt[,.SD[which.max(abs(V1))], by = c('tool')]
dt$tool <- toupper(gsub("_factors", "", dt$tool))
p6 <- ggplot(dt, aes(x = tool, y = abs(V1))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  coord_flip() + 
  labs(y = 'Correlation with Tumor Mutation Burden') + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5)) 
# immune phenotypes 
dt <- do.call(rbind, lapply(names(LFs_imtrp), function(tool) {
  M <- LFs_imtrp[[tool]]$mariathasan.gex
  # find LFs predictive of immune phenotypes 
  dt <- get_factor_specific_variables(M, colData$`Immune phenotype`)[order(padj)]
  dt$tool <- tool
  return(dt)
}))

dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')]
dt$tool <- toupper(gsub("_factors", "", dt$tool))
p7 <- ggplot(dt[,.SD[which.min(padj)],by = c('ref_cl', 'tool')], 
             aes(x = ref_cl, y = -log10(padj))) + 
  geom_bar(stat = 'identity', aes(fill = tool), position = 'dodge') +
  geom_hline(yintercept = -log10(0.05)) + 
  labs(x = 'Tumor Immune Phenotypes') + 
  coord_flip() + 
  theme(legend.title = element_blank(), axis.title.y = element_blank(), 
        plot.title = element_text(hjust = 0.5))

Fig <- cowplot::plot_grid(
  cowplot::plot_grid(p1, p6, p7, nrow = 1, labels = 'AUTO'),
  cowplot::plot_grid(p2$gtable, p3$plot, p4, p5, nrow = 2, labels = c('D', 'E', 'F', 'G')),
  nrow = 2, 
  rel_heights = c(1,2)
  ) 
fig_count <- fig_count + 1
Figures[[fig_count]] <- Fig
fig.name <- paste0('Figure', fig_count, '.immunotherapy_response.pdf')
names(Figures)[fig_count] <- fig.name
ggsave(filename = file.path(FigureFolder, fig.name), 
       plot = Fig, units = 'in', 
       width = 12, height = 14)
Legends[[fig_count]] <- "Predicting responders to anti-PD-L1 treatment in a cohort of Metastatic Urothelial
Cancers (N=348) using MAUI. A) AUC curve for predicting Complete/Partial Responders (CR/PR) to the anti-PD-L1
treatment using an Elastic Net Model on a subset of MAUI latent factors individually predictive of treatment response groups. B) Average activation score of top latent factors predictive of treatment response groups. C) Top pathways enriched among top 100 genes that contribute to the top latent factor (LF44). D) Survival stratification of the 
cohort based activation score of LF44 (high = LF44 > 1.4, low = LF44 < 1.4). E) Top 10 genes contributing to LF44
F) Correlation between LF44 activation score and the tumor (neo-antigen) mutation burden"
```
